---
title: 'GSE19711'
output: html_document
---

# Authors


- Rimvydas Noreika
- Tautminas Cibulskis
- Gabrielius Erignis
- Rolandas Porėjus


# Analysis
 ```{r, echo=FALSE, include=FALSE}
library(GEOquery)
 ```
- Automatically download the data from GEO
```{r}
gset = getGEO("GSE19711", destdir="./")
```
- Obtain the matrix of beta values where each row corresponds to probes and each column corresponds to samples
```{r}
gsetexpr = exprs(gset[[1]])
rownames(gsetexpr)[1:5]
colnames(gsetexpr)[1:5]
```
- How many samples and how many probes do you have in the data?
```{r}
sampleset = colnames(gsetexpr)
length(samples)

probeset = rownames(gsetexpr)
length(probeset)
```
- How are the beta values distributed?

 ```{r}
 hist(gsetexpr)
 ```
 
- Do your probes have names?
```{r}
probeset[1:5]
```
- Do you have annotation that tells the coordinate (in hg19) of each probe and its genomic features (such as related gene name)?
```{r}
annotation = getGEO("GPL8490", destdir = "./")
annotation = Table(annotation)
head(annotation)
```
- Do you know which samples correspond to healthy individuals, and which samples correspond to the sick ones?
```{r}
disease = pData(phenoData(gset[[1]]))
disease[1:20,1]
```
For each probe compute a t-test to verify if the distributions of beta values within the probe significantly differ between the two groups. From the t-test, obtain the p value.
```{r}
pvalarray = array()
for (i in 1:length(probeset)){
  case = gsetexpr[i, grep("Case", disease$`sample type:ch1`)]
  control = gsetexpr[i, grep("Control", disease$`sample type:ch1`)]
  tt = t.test(case, control)
  pvalarray[i] = tt$p.value
} 
```

Plot the distribution of p values. What is the expected distribution? How does it differ from what you get?
We get Anti-conservative p-values. All null p-values form a flat distribution. Peak is close to 0 - that is where the alternative hypotheses are. Not all p-values less than .05 are significant. High p-values indicate false negatives - hypotheses that are not detectable with our test. The left of the histogram shows significance. The right side shows how many p-values are null.

```{r}
hist(pvalarray, breaks = 100)
```

Performance-wise, how long will it take to compute the test for all probes?
```{r}
system.time( for (i in 1:length(probeset)){
  case = gsetexpr[i, grep("Case", disease$`sample type:ch1`)]
  control = gsetexpr[i, grep("Control", disease$`sample type:ch1`)]
  tt = t.test(case, control)
  pvalarray[i] = tt$p.value
})
```

