---
title: 'GSE19711'
output: html_document
---

# Authors


- Rimvydas Noreika

- Tautminas Cibulskis

- Gabrielius Erignis


# Analysis
 ```{r, echo=FALSE, include=FALSE}
library(GEOquery)
 ```
- Automatically download the data from GEO
```{r}
gset = getGEO("GSE19711", destdir="./")
```
- Obtain the matrix of beta values where each row corresponds to probes and each column corresponds to samples
```{r}
gsetexpr = exprs(gset[[1]])
rownames(gsetexpr)[1:5]
colnames(gsetexpr)[1:5]
```
- How many samples and how many probes do you have in the data?
```{r}
sampleset = colnames(gsetexpr)
length(samples)

probeset = rownames(gsetexpr)
length(probeset)
```
- How are the beta values distributed?
 ```{r}
 hist(gsetexpr)
 ```
- Do your probes have names?
```{r}
probeset[1:5]
```
- Do you have annotation that tells the coordinate (in hg19) of each probe and its genomic features (such as related gene name)?
```{r}
annotation = getGEO("GPL8490", destdir = "./")
annotation = Table(annotation)
head(annotation)
```
- Do you know which samples correspond to healthy individuals, and which samples correspond to the sick ones?
```{r}
disease = pData(phenoData(gset[[1]]))
disease[1:20,1]
```
