---
title: 'GSE19711'
output: html_document
---

# Authors


- Rimvydas Noreika

- Tautminas Cibulskis

- Gabrielius Erignis


# Analysis
 ```{r}
 library(GEOquery)
 ```
- Automatically download the data from GEO
```{r}
# gset = getGEO("GSE19711", GSEMatrix=FALSE, getGPL=FALSE)
```
- Obtain the matrix of beta values where each row corresponds to probes and each column corresponds to samples
```{r}
gsmplatforms = lapply(GSMList(gset),function(x) {Meta(x)$platform_id})
gsmlist = Filter(function(gsm) {Meta(gsm)$platform_id=='GPL8490'},GSMList(gset))
probesets = Table(GPLList(gset)[[1]])$ID
data.matrix <- do.call('cbind',
                lapply(
                       gsmlist,function(x) 
                        {
                         tab <- Table(x)
                         mymatch <- match(probesets,tab$ID_REF)
                         return(tab$VALUE[mymatch])
                        }
                       )
                      )
data.matrix <- apply(data.matrix,2,function(x) {as.numeric(as.character(x))})
data.matrix[1:5,]
```
- How many samples and how many probes do you have in the data?
```{r}
samples = names(GSMList(gset))
lenght(samples)

probesets = Table(GSMList(gset)[[1]])$ID
lenght(probesets)
```
- How are the beta values distributed?
 ```{r}
 betaVals = Table(GSMList(gset)[[1]])$VALUE
 
 ```
- Do your probes have names?
```{r}
probs = Table(GSMList(gset)[[1]])$ID
probs[1:5]
# Taip
```
- Do you have annotation that tells the coordinate (in hg19) of each probe and its genomic features (such as related gene name)?

- Do you know which samples correspond to healthy individuals, and which samples correspond to the sick ones?
