---
title: "GSE30229"
date: '2018 m. kovas'
output: html_document
---
# Authors
- Edgaras Legus
- Vitalija Misiukonytė
- Brigita Ižganaitytė
- Gabrielė Dilytė
- Dalia Masilionytė
- Reda Vaisėtaitė
- Paulius Matijošaitis
- Dovilė Patiejūnaitė

# Analysis
## Step 2
Downloading the data using GEO 
```{r, include=TRUE, echo=TRUE}
library(GEOquery)
GSE30229 <- getGEO("GSE30229", destdir = "./")
```

Obtaining the betaValueMatrix of beta values where each row corresponds to probes (names) and each column corresponds to samples

```{r, include=TRUE, echo=TRUE}
betaValueMatrix <- exprs(GSE30229[[1]])
head(rownames(betaValueMatrix))
head(colnames(betaValueMatrix))
```

Counting samples and probes in our data
```{r, include=TRUE, echo=TRUE}
samples <- colnames(betaValueMatrix)
probes <- rownames(betaValueMatrix)
length(samples)
length(probes)
```

Distribution of beta values
```{r, include=TRUE, echo=TRUE}
hist(betaValueMatrix, breaks=1000, xlab = "Beta Value ", border = "green", main = "Beta value distribution")
```

Names of probes
```{r, include=TRUE, echo=TRUE}
head(rownames(betaValueMatrix))
```

Annotation that tells the coordinate (in hg19) of each probe and its genomic features
```{r, include=TRUE, echo=TRUE}
annotate <- getGEO("GPL8490", destdir = "./")
annotate <- Table(annotate)

annotated_probes <- intersect(annotate$ID, rownames(betaValueMatrix))
cat("Probes number: ", length(annotated_probes), "\n")

i <- match(annotated_probes, annotate$ID)
annotate <- annotate[i, ]

i <- match(annotated_probes, rownames(betaValueMatrix))
betaValueMatrix <- betaValueMatrix[i,]
stopifnot(all(rownames(betaValueMatrix) == annotate$ID))

head(annotate)
```

Samples which correspond to healthy individuals, and which samples correspond to the sick ones
```{r, include=TRUE, echo=TRUE}
sickness <- pData(phenoData(GSE30229[[1]]))
sickness[1:20,1]
```

## Step 3

For each probe compute a t-test to verify if the distributions of beta values within the probe significantly differ between the two groups.

```{r, include=TRUE, echo=TRUE}
controlSamples <- which(sickness$`disease:ch1` == 'control')
caseSamples <- which(sickness$`disease:ch1` == 'HNSCC case')
t.test(betaValueMatrix[1,controlSamples], betaValueMatrix[1,caseSamples])
```

From the t-test, obtain the p value.
```{r, include=TRUE, echo=TRUE}
pvalues = c()
for (i in 1:length(rownames(betaValueMatrix))) {
     pvalues[i] <- t.test(betaValueMatrix[i,controlSamples], betaValueMatrix[i,caseSamples])$p.value
     
}
```
Plot the distribution of p values. What is the expected distribution? How dows it differ from what you get?

```{r, include=TRUE, echo=TRUE}
#The peak close to 0 is tall, so there are many p-values close to 0 which means that there is a lot of significant values.
#The “depth” of the histogram on the right side shows the values that are null.
hist(pvalues, col = "green", breaks = 100)
```

Performance-wise, how long will it take to compute the test for all probes?
```{r, include=TRUE, echo=TRUE}
# We used system.time() function to measure the performance. It took 30.100 seconds to perform all the t-tests
```

