---
title: 'GSE50409'
output: html_document
---
# Authors

- Leonas Rėčkus
- Liza Tofan
- Aurimas Vilys
- Karolina Butrimaitė
- Ignas Labalaukis
- Domas Motiejūnas

# Analysis

```{r setup, echo=FALSE, include=FALSE}
library(GEOquery)
library(data.table)
```

- Automatically download the data from GEO

```{r echo=FALSE, include=FALSE}
gse50409 <- getGEO("gse50409", destdir="./")
```

- Obtain the matrix of beta values where each row corresponds to probes and each column corresponds to samples

```{r}
matrix <- exprs(gse50409[[1]])
head(rownames(matrix))
head(colnames(matrix))
```

- How many samples and how many probes do you have in the data?

Number of probes:
```{r}
probes <- rownames(matrix)
length(probes)
```

Number of samples: 
```{r}
samplesNames <- colnames(matrix)
length(samplesNames)
```

- How are the beta values distributed?

```{r out.width = '100%'}
hist(matrix, breaks=1000)
```

- Do your probes have names?

These are the probe names:
```{r}
head(rownames(matrix))
```

- Do you have annotation that tells the coordinate (in hg19) of each probe and its genomic features (such as related gene name)?

```{r}
annotation <- getGEO("GPL13534", destdir = "./")
annotation <- Table(annotation)
# setDT(annotation)

# Now match the rows in annotation that are present in our data
commonProbes <- intersect(annotation$ID, rownames(matrix))
cat("Number of probes for which hg19 annotation is available: ", length(commonProbes), "\n")

i <- match(commonProbes, annotation$ID)
annotation <- annotation[i, ]

i <- match(commonProbes, rownames(matrix))
matrix <- matrix[i,]
stopifnot(all(rownames(matrix) == annotation$ID))

head(annotation)
```

- Do you know which samples correspond to healthy individuals, and which samples correspond to the sick ones?

```{r}
disease <- pData(phenoData(gse50409[[1]]))
disease[1:20,1]
```
