---
title: 'GSE50409'
output: html_document
---
# Authors

- Leonas Rėčkus
- Liza Tofan
- Aurimas Vilys
- Karolina Butrimaitė
- Ignas Labalaukis
- Domas Motiejūnas

# Analysis
```{r setup, echo=FALSE, include=FALSE}
library(GEOquery)
```
- Automatically download the data from GEO
```{r echo=FALSE, include=FALSE}
gse50409 <- getGEO("gse50409", destdir="./")
```
- Obtain the matrix of beta values where each row corresponds to probes and each column corresponds to samples
```{r}
matrix <- exprs(gse50409[[1]])
matrix[1:20, 1:20]
```
- How many samples and how many probes do you have in the data?
```{r}
probes <- featureNames(featureData(gse50409[[1]]))
length(probes)
samplesNames <- sampleNames(phenoData(gse50409[[1]]))
length(samplesNames)
```
- How are the beta values distributed?
```{r out.width = '100%'}
boxplot(matrix, outline = FALSE, boxwex = 0.7, notch = T, las = 2, main = "GSE50409")
```
- Do your probes have names?
```{r}
pData(featureData(gse50409[[1]]))[1:20,1:2]
t1 <- row.names(featureData(gse50409[[1]]))
t2 <- as.character(featureData(gse50409[[1]])$Name)
identical(t1, t2)
#Probe names are identical to probe ID
```
- Do you have annotation that tells the coordinate (in hg19) of each probe and its genomic features (such as related gene name)?
```{r}
gse50409
```
- Do you know which samples correspond to healthy individuals, and which samples correspond to the sick ones?
```{r}
disease <- pData(phenoData(gse50409[[1]]))
disease[1:20,1]
```